---
title: "Simulation Studies of Test Equating"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(MplusAutomation)
require(mirt)
require(tidyr)
require(ggplot2)
require(psych)
source("~/Research/Code/runMplus.R")
source("~/Research/Code/extractVarNames.R")
source("../Code/simulation_functions.R")
```

### Technical Decision Points

- Simulation platform
- Platform to estimate ability from simulated datasets
  - Estimation method
- Metric for comparing different simulation scenarios

### Simulation Sanity Check

- Should be able to recreate ability metric of simulated item-level dataset in a single sample 
  - Flat Bland-Altmann plot - (estimated ability - true ability)/true ability
    - Dispersion around 0 line consistent with expected standard error of measurement
  - Simulation parameters
    - 30 dichotomous items
    - item difficulty equaly spaced from -3 to +3
    - item discrimination all 2.0
    - sample size = 500
    - theta simulated from N(0,1)
    
```{r message=FALSE, echo=FALSE, warning=FALSE, include=FALSE, cache=TRUE}

### Simulate item level dataset

discr <- c(2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2)

set.seed(4357)
ds1 <- sim.poly.npl(nvar = 30 ,n = 500,low=-3,high=3,a=discr,
        c=0,z=1,d=NULL, mu=0,sd=1,cat=2)

diff <-  data.frame(ds1[["difficulty"]])
names(diff) <- "difficulty"
disc <-  data.frame(ds1[["discrimination"]])
names(disc) <- "discrimination"


t1 <- data.frame(ds1[["items"]])
t2 <- data.frame(ds1[["theta"]])

names(t2) <- "ability"

t2 <- cbind(t2,t1)


### Mplus code for simulation sanity check

title <- "Simulated IRT Analysis"

analysis <- "estimator = mlr;
"

output <- "sampstat;stdyx;svalues;"

varsim2 <- c('V1','V2','V3','V4','V5','V6','V7','V8','V9','V10',
            'V11','V12','V13','V14','V15','V16','V17','V18','V19','V20',
            'V21','V22','V23','V24','V25','V26','V27','V28','V29','V30')

varcatsim2 <- "
categorical = v1-v30;
"

savedata <- "
FILE = 'irt_sim_fscores_3.dat';
SAVE = FSCORES;
"

model_sim_4 <- "
cog by v1-v30@2;
"

for(i in 1:30){
  if (i==1){
    thresh <- paste("[",varsim2[1],"$1","@",diff[1,1]*disc[1,1],"];\n",sep="")
  } else {
    thresh <- paste(thresh,"[",varsim2[i],"$1","@",diff[i,1]*disc[i,1],"];\n",sep="")
  }
}

model_sim_4 <- paste(model_sim_4,thresh,sep="\n")

# Mplus EAP factor scores

rdata <- t2
modelout <- "../Analysis/irt_sim_mplus_eap.inp"

mplobj <- mplusObject(TITLE = title, MODEL = model_sim_4, VARIABLE = varcatsim2, 
        ANALYSIS = analysis, OUTPUT = output, SAVEDATA = savedata, 
        rdata = rdata, usevariables = varsim2)
res_irt_mplus_eap <- mplusModeler(mplobj,modelout=modelout,run=1,check=TRUE)


fsc <- readModels("../Analysis/irt_sim_mplus_eap.out",what="savedata")$savedata

t4 <- cbind(t2,fsc[,c("COG","COG_SE")])

t4$resid <- t4$COG - t4$ability

# Mplus plausible values factor scores

savedata2 <- "
FILE = 'irt_sim_fscores_pv.dat';
SAVE = FSCORES(50,10);
"
modelout <- "../Analysis/irt_sim_mplus_pv.inp"

analysis2 <- sub("mlr","bayes",analysis)

mplobj2 <- mplusObject(TITLE = title, MODEL = model_sim_4, VARIABLE = varcatsim2, 
        ANALYSIS = analysis2, OUTPUT = output, SAVEDATA = savedata2, 
        rdata = rdata, usevariables = varsim2)
res_irt_mplus_pv <- mplusModeler(mplobj2,modelout=modelout,run=1,check=TRUE)


# res_irt_sim_4 <- runMplus(TITLE = title, MODEL = model_sim_4, usevariables = varsim2,
#                           rdata=rdata, OUTPUT = output, modelout = modelout,
#                           VARIABLE = varcatsim2, ANALYSIS=analysis2, delete_data=FALSE,
#                           SAVEDATA=savedata2)
rm(res_irt_mplus_pv)

fsc <- readModels("../Analysis/irt_sim_mplus_pv.out",what="savedata")$savedata

t5 <- cbind(t2,fsc[,c("COG.Mean","COG.Median","COG.Standard.Deviation")])

t5$resid <- t5$COG.Median - t5$ability



### mirt simulation and estimation

theta1 <- rnorm(500,0,1)
dataset1 <- simdata(disc, diff, 500, Theta=as.matrix(theta1),  itemtype = '2PL')
dataset1 <- data.frame(cbind(theta1,dataset1))
names(dataset1) <- sub("theta1","ability",names(dataset1))


mdl <- mirt.model('cog = 1-30')


mcal_par <- mirt(dataset1[,2:31],mdl,pars='values')

mcal_par[mcal_par$name == "a1","value"] <- 2.0
mcal_par[mcal_par$name == "a1","est"] <- FALSE
mcal_par[mcal_par$name == "d","value"] <- diff
mcal_par[mcal_par$name == "d","est"] <- FALSE
mcal_par[mcal_par$item == "GROUP","est"] <- TRUE


mcal <- mirt(dataset1[,2:31],mdl,itemtype='2PL',pars=mcal_par)

# fsc <- fscores(mcal,full.scores=TRUE, plausible.type="MH",plausible.draws=50)
# fsc_med <- data.frame(list(fsc[[5]],fsc[[10]],fsc[[15]],fsc[[20]],fsc[[25]],fsc[[30]],fsc[[35]],
#     fsc[[40]],fsc[[45]],fsc[[50]]))
# names(fsc_med) <- c("D1","D2","D3","D4","D5","D6","D7","D8","D9","D10")
# fsc_med$ability_est <- apply(fsc_med,1,median)

fsc <- data.frame(fscores(mcal,full.scores=TRUE, method="plausible"))
names(fsc) <- "ability_est"
t6 <- cbind(dataset1,fsc)

t6$resid <- t6$ability_est - t6$ability



fsc <- data.frame(fscores(mcal,full.scores=TRUE, method="EAP"))
names(fsc) <- "ability_est"
t7 <- cbind(dataset1,fsc)


t7$resid <- t7$ability_est - t7$ability


fsc <- data.frame(fscores(mcal,full.scores=TRUE, method="ML"))
names(fsc) <- "ability_est"
t8 <- cbind(dataset1,fsc)

t8$ability_est <- ifelse(t8$ability_est %in% c(-Inf,Inf),NA,
    t8$ability_est)


t8$resid <- t8$ability_est - t8$ability





```

```{r message=FALSE, echo=FALSE, warning=FALSE, include=FALSE}

item_list_1 <- c("Item_1","Item_2","Item_3","Item_4","Item_5","Item_6","Item_7",
                 "Item_8","Item_9","Item_10","Item_11","Item_12","Item_13","Item_14",
                 "Item_15","Item_16","Item_17","Item_18","Item_19","Item_20","Item_21",
                 "Item_22","Item_23","Item_24","Item_25","Item_26",
                 "Item_27","Item_28","Item_28","Item_29","Item_30")
item_list_2 <- c("Item_1","Item_2","Item_3","Item_4","Item_5","Item_6","Item_7",
                 "Item_8","Item_9","Item_10","Item_11","Item_12","Item_13","Item_14",
                 "Item_15","Item_16","Item_17","Item_18","Item_19","Item_20","Item_21",
                 "Item_22","Item_23","Item_24","Item_25","Item_26",
                 "Item_27","Item_28","Item_28","Item_29","Item_30")


tinfo <- infoSim(seed=21589,n_group=2,grp_mean=c(0.5,-0.5),grp_sd=c(1,1),
    n_samp=500,pars=mcal_par,itms1=item_list_1,itms2=item_list_2)

tinfo2 <- tinfo %>% gather(key,value,c("info_1","info_2"))
tinfo2$key <- ifelse(tinfo2$key %in% "info_1","Group 1","Group 2")

```

### Test Information of Simulated Sanity Check Test

```{r message=FALSE, echo=FALSE, warning=FALSE}

ggplot(data=tinfo2,aes(x=ability, y=value, color=key)) + geom_line(lwd=1) + xlab("Ability") +
  ylab("Information")

```

 
### Sanity Check 1 - R psych simulation, ability estimated with Mplus, factor score method = EAP
ability_est mean = `r mean(t4$COG)`   
ability_est sd = `r sd(t4$COG)`   
rmse = `r sqrt(mean(t4$resid^2))`   

```{r message=FALSE, echo=FALSE, warning=FALSE}
# plot(t4$ability,t4$COG)
# plot(t4$ability,t4$resid)
# abline(lm(t4$resid ~ t4$ability))
# 

ggplot(data=t4,aes(x=ability,y=resid)) + geom_point() + geom_smooth(method="lm")

```

### Sanity Check 2 - R psych simulation, ability estimated with Mplus, factor score method = plausible values
ability_est mean = `r mean(t5$COG.Median)`   
ability_est sd = `r sd(t5$COG.Median)`   
rmse = `r sqrt(mean(t5$resid^2))`   

```{r message=FALSE, echo=FALSE, warning=FALSE}
# plot(t4$ability,t4$COG)
# plot(t4$ability,t4$resid)
# abline(lm(t4$resid ~ t4$ability))
# 

ggplot(data=t5,aes(x=ability,y=resid)) + geom_point() + geom_smooth(method="lm")

```

### Sanity Check 3 - R mirt simulation and estimation of ability, factor score method = plausible values
ability_est mean = `r mean(t6$ability_est)`   
ability_est sd = `r sd(t6$ability_est)`   
rmse = `r sqrt(mean(t6$resid^2))`   

```{r message=FALSE, echo=FALSE, warning=FALSE}
# plot(t4$ability,t4$COG)
# plot(t4$ability,t4$resid)
# abline(lm(t4$resid ~ t4$ability))
# 

ggplot(data=t6,aes(x=ability,y=resid)) + geom_point() + geom_smooth(method="lm")

```

### Sanity Check 4 - R mirt simulation and estimation of ability, factor score method = EAP
ability_est mean = `r mean(t7$ability_est)`   
ability_est sd = `r sd(t7$ability_est)`   
rmse = `r sqrt(mean(t7$resid^2))`   

```{r message=FALSE, echo=FALSE, warning=FALSE}
# plot(t4$ability,t4$COG)
# plot(t4$ability,t4$resid)
# abline(lm(t4$resid ~ t4$ability))
# 

ggplot(data=t7,aes(x=ability,y=resid)) + geom_point() + geom_smooth(method="lm")

```
  
### Sanity Check 5 - R mirt simulation and estimation of ability, factor score method = ML
ability_est mean = `r mean(t8$ability_est,na.rm=TRUE)`   
ability_est sd = `r sd(t8$ability_est,na.rm=TRUE)`   
rmse = `r sqrt(mean(t8$resid^2,na.rm=TRUE))`   

```{r message=FALSE, echo=FALSE, warning=FALSE}
# plot(t4$ability,t4$COG)
# plot(t4$ability,t4$resid)
# abline(lm(t4$resid ~ t4$ability))
# 

ggplot(data=t8,aes(x=ability,y=resid)) + geom_point() + geom_smooth(method="lm")

```

### Recommendations

- R mirt for simulation and estimation
  - Common platform for both components
  - Accomodates dichotomous and polytomous items
- RMSE as metric for comparing scenarios
- Ability score estimation method has tradeoffs
  - plausible values best in reproducing true ability metric
    - greater dispersion around true ability (RMSE)
  - EAP has low dispersion around true ability
    - more shrinkage of estimated ability metric, though may be acceptable
  - ML reproduces metric and has low dispersion around true ability
    - estimated abilityis not defined for cases with all correct or all incorrect responses
  -Should we consider transforming estimated ability scale to 0,1 metric?
  
 
 
```{r message=FALSE, echo=FALSE, warning=FALSE, include=FALSE, cache=TRUE}
 
 ### mirt calibration of tics

# input PITCH tics dataset
tics <- read.csv("../Data/PITCH-response-data-fhl.csv")

# HRS items
vars <- c('UBAK','UDAT','UDAY','UDWR','UIWR','UMON','UNM1','UNM2','UNM5','UNM6',
     'USUB','UYER')
# MHAS items
varsm <- c("UDAY","UFCO2","UFRE1","UMON","UVSC","UWD","UWR1","UWR2","UWR3","UYER")

# select HRS items and persons
hrs <- tics[tics$study_name_short %in% c("HRS_CODA_W6","HRS_W6"),vars]
hrs$n_itm <- apply(hrs,1,function(x) sum(!is.na(x)))
hrs <- hrs[!hrs$n_itm ==0,]

m1hrs <- mirt.model('cog = 1-12') # creates mirt model object
hrs_par <- mirt(hrs[,1:12],m1hrs,pars='values') # generates item parameters file that can be edited to guide further analyses

hrscal <- mirt(hrs[,1:12],m1hrs,pars=hrs_par) # IRT calibration
# coef(hrscal)

m2hrs <- mirt.model('cog = 1-11')

# HRS calibration excluding delayed word recall
hrs_par_0dr <- mirt(hrs[,c(1:3,5:12)],m2hrs,pars='values')
hrscal_0dr <- mirt(hrs[,c(1:3,5:12)],m2hrs,pars=hrs_par_0dr)

# HRS calibration excluding immediate word recall
hrs_par_0ir <- mirt(hrs[,c(1:4,6:12)],m2hrs,pars='values',technical=list(removeEmptyRows=TRUE))
hrscal_0ir <- mirt(hrs[,c(1:4,6:12)],m2hrs,pars=hrs_par_0ir,technical=list(removeEmptyRows=TRUE))
# coef(hrscal_0ir)

#  mirt calibration of MHAS 

mex <- tics[tics$study_name_short %in% c("MHAS_W1","MHAS_W2"),varsm]
mex$n_itm <- apply(mex,1,function(x) sum(!is.na(x)))
mex <- mex[!mex$n_itm ==0,]

m1mex <- mirt.model('cog = 1-10')
mex_par <- mirt(mex[,1:10],m1mex,pars='values')

mexcal <- mirt(mex[,1:10],m1mex,pars=mex_par)
# coef(mexcal)

#  mirt calibration of combined HRS and MHAS 

hrme <- tics[tics$study_name_short %in% c("HRS_CODA_W6","HRS_W6") |
    tics$study_name_short %in% c("MHAS_W1","MHAS_W2"),union(vars,varsm)]
hrme$n_itm <- apply(hrme,1,function(x) sum(!is.na(x)))
hrme <- hrme[!hrme$n_itm == 0,]

m1hrme <- mirt.model('cog = 1-19')
hrme_par <- mirt(hrme[,1:19],m1hrme,pars='values')

hrmecal <- mirt(hrme[,1:19],m1hrme,pars=hrme_par)
# coef(hrmecal)


```

### HRS TICS Information (immediate and delayed word list recall included)
info_all - all items including immediated word recall and delayed word recall are included   
info_no_dr - all items except delayed word recall are included   
info_no_ir - all items except immediated word recall are included   

```{r message=FALSE, echo=FALSE, warning=FALSE}

info1 <- infoCalc(hrscal)
info2 <- infoCalc(hrscal_0dr)
info3 <- infoCalc(hrscal_0ir)

info <- merge(info1,info2,by="ability")
info <- merge(info,info3,by="ability")
names(info) <- c("ability","info_all","info_no_dr","info_no_ir")
inf_comb <- info %>% gather(key,information,c("info_all","info_no_dr","info_no_ir"))

ggplot(data=inf_comb,aes(x=ability,y=information,col=key)) + geom_line(lwd=1)

```

Note. Information values for all items included in model over estimate true information values due to unidimensionality and local independence assumptions not being met in this model.

### MHAS TICS Information (all items included)

```{r message=FALSE, echo=FALSE, warning=FALSE}

info1 <- infoCalc(mexcal)

ggplot(data=info1,aes(x=ability,y=information)) + geom_line(lwd=1)

```

Note. Information values for all items included in model over estimate true information values due to unidimensionality and local independence assumptions not being met in this model.

### Combined HRS and MHAS TICS Information (all items included)

```{r message=FALSE, echo=FALSE, warning=FALSE}

info1 <- infoCalc(hrmecal)

ggplot(data=info1,aes(x=ability,y=information)) + geom_line(lwd=1)

```

Note. Information values for all items included in model over estimate true information values due to unidimensionality and local independence assumptions not being met in this model.

### HRS TICS Item Parameters (all items included)

```{r message=FALSE, echo=FALSE, warning=FALSE}
print(coef(hrscal))
```


### MHAS TICS Item Parameters (all items included)

```{r message=FALSE, echo=FALSE, warning=FALSE}
print(coef(mexcal))
```

### Combined HRS and MHAS TICS Item Parameters (all items included)

```{r message=FALSE, echo=FALSE, warning=FALSE}
print(coef(hrmecal))
```


### Simulate item response dataset using item parameters from combined HRS and MHAS data

```{r message=FALSE, echo=TRUE, warning=FALSE}
# This chunk show code to create one simulated item response dataset based on empirically
# estimated parameters from and IRT analysis of the combined HRS and MHAS samples
theta1 <- data.frame(rnorm(500,0.5,1))
  names(theta1) <- "theta1"
  theta1$group <- "HRS"
  theta2 <- data.frame(rnorm(500,-0.5,1))
  names(theta2) <- "theta1"
  theta2$group <- "MHAS"
  theta1 <- rbind(theta1,theta2)
  

  ds2 <- data.frame(simdata(N=1000, model=hrmecal, Theta=as.matrix(theta1$theta1)))
  
  ds2 <- cbind(theta1,ds2)
 
  m1hrme <- mirt.model('cog = 1-19')
  hm_par <- mirt(ds2[,3:21],m1hrme,pars='values')

  hmcal <- mirt(ds2[,3:21],m1hrme,pars=hm_par)
  # coef(hmcal)
 
```
